Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-15 paper.tex
added in Aarons latest edits

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: The
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce the
  new therapeutic molecules.  Only within the last decade -- and as an
  effect of a change in mindsets, government-mandated policy changes
  and last but not least because of chemists volunteering their time
  for an Open Science ``movement'' -- have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Making chemical information public is important because the drugs we
take are the result of intensive and costly research, which is
facilitated by the amount of information available. The more
information is shared (a process very hard to achieve in
pharmaceutical companies), the more information is available to each
researcher, facilitating the development of novel treatments. The
relevance of computer science to this research becomes clear when we
consider the quantity and types of data available -- currently a
single data base such as PubChem contains more than 34 million
chemical structure records, along with an even larger number of
annotations such as synonyms, known targets, mode of action,
regulatory approval status and so on.  Simply considering the synonym
list, PubChem stores 50.1 million synonym for 19.6 million compounds.

@@

Cheminformatics comprises different areas, which broadly speaking can
be divided into three fields: \emph{capturing data} (using lab
notebooks or potentially using formats such as the Chemical Markup
Language for publications); \emph{storing data} (designing database
schemas, devising ontologies) and \emph{mining data} (such as for
bioactivity prediction, which might involve algorithms based on
graphs).  However, the type of information being analyzed is different
from biological information: while bioinformatics often deals with
sequences, the domain of cheminformatics is chemical structures. In
the case of the former, information can very frequently be represented
as one-dimensional strings, which are relatively easy to handle
computationally. In the case of the latter however, chemical
structures may possess rings, branches as well as multiple valid
representations of the same structure (such as so-called
``tautomers'', where hydrogen atoms can be positioned in different
places of the molecules), potentially giving rise to
ambiguities. Hence, chemical structures are often reported to be more
difficult to standardize.

@@

Although graphs implicitly encodes the 3D structure of a molecule
(when combined with knowledge about favored bond angles and
distances), many different low energy 3D structures, or conformers,
may be consistent with the same graph structure. In addition, the 3D
structure may have a ``secondary'' geometrical arrangement of features
(such as the presence of a right-handed helix) which cannnot be
encoded in the graph. Thus we have 3D representations  
(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
goes further and represents a molecular surface, encoding some
property (such as lipophilicity) that varies over the surface.

@@

There is considerable progress in open chemical and bioactivity
databases like ChEMBL and NCBI-PubChem, which are both freely
available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Still, the integration of
chemical databases with each other remains a challenge, not only due
to the normalization problems in the chemical and bioactivity
landscape, but also due to sheer data volume. One problem is the
annotation with the ``protein targets'' a molecule binds to -- here
dozens of protein and gene identifiers exist, generally without the
availability of a one-to-one mapping from each ontology, which makes
merging databases often a cumbersome exercise.

Given the requirement to ``make sense'' of as much chemical data at
hand as possible, we have to be able to search in complex data, for
example multi-labelled molecular graphs, where graph labels can be
different. Hence, we need to be able to capture molecular graphs in a
risk-bounded machine encoded way.

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. When one considers the need to include other aspects such as
protonation and tautomer states, the challenge increases. The need for
a unique representation that is invariant with respect to the atom
ordering arises due to the fact that graph isomorphism (i.e., checking
whether two molecules are the same) is an expensive operation. The
first such canonicalization algorithm was described by Morgan
\cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format \cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr (Warr \cite{Warr:2011vn}) for an extensive dicussion on chemical
structure representations.

With each database using their own algorithm for molecular encoding,
automated exchange of data between different databases was hindered.
As more and more data became available online, the use of unique
structure-based identifiers became a pressing need, resulting in the
IUPAC International Chemical Identifier (InChI) to be developed. The
InChI is a non-proprietary, structured textual identifier for chemical
entities~\cite{inchi}. InChI identifiers are not intended to be read
and understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. (Unfortunately, after implementation and publication of
the hashing scheme it became apparent that in particular in case of
stereochemically related molecules clashes of the InChI key take place
in some, albeit rare cases.) Figure~\ref{figure:smiles} illustrates
the SMILES, InChI and InChIKey for lipoic acid.

@@

The InChI is now widely used for matching identical chemical
structures, but it is still limited. For example, it cannot
differentiate between certain types of stereoisomers:
Figure~\ref{figure:cistrans} (in the Appendix) illustrates two
stereoisomers for which the generated InChI is the same.

@@

Graph substructure matching is a variant of \emph{graph isomorphism},
which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is infeasible \cite{Weininger:2011ly}. Speedups can be
obtained via the use of heuristics such as structural
\emph{fingerprints}. Fingerprints encode characteristic features of a
given chemical structure, usually in a fixed-length
bitmap. Fingerprints fall broadly into two categories: structure keys
and hashed keys. In the former, each bit position corresponds to a
distinct substructure such as an alcohol or keto group. Examples
include MACCS and PubChem keys. The latter class are designed such
that substructural patterns are represented as strings and then hashed
to a random bit position. As a result, a given position can encode
multiple substructures. The advantage of such fingerprints is that
they can cover an arbitrarily large collection of substructures such
as paths of length $N$ or circular environments and so on. Examples
include ECFPs and Daylight fingerprints.

Given a binary fingerprint, we can first ``pre-screen'' a database, to
ignore molecules that cannot possibly match the query. This is
achieved by requiring that all bits in a query fingerprint must also
be present in the target fingerprint. Since the target fingerprints
are pre-computed, comparing binary strings can be performed extremely
rapidly on modern hardware. As a result, we apply the actual
isomoprhism test only on those molecules that pass the
screen. Fingerprints can also be used to rapidly search databases for
similar molecules, in which case a similarity metric such as the
Tanimoto coefficient is used to compare the query and target
fingerprints. While fast, heuristics have been developed
\cite{Swamidass:2007ve} to further
speed up similarity searches.

@@

The preceding section focuses on the choice of representation of
chemical structures. With the profusion of databases that contain
details of small molecules, an emerging problem is the integration of
such databases. Such integration forms the basis of ``systems'' level
investigations of biological systems. A current challege in
cheminformatics is how to identify ``equivalent'' objects across data
sources. At one level, equivalent compounds can be identified via
canonical structure representations (InChI). But there are other
equivalences beyond structure. For example, compounds may bind to the
same or similar targets; they may be have similar biological function
(say, anti-neoplastic) and so on. To assess such higher level
characteristics, ontologies and semantic technologies have been
employed, to first, support annotations of chemical systems and
second, enable semantic analyses of chemical systems in conjunction
with other biological systems (such as pathways). A number of
challenges remain in this area - the use of ontologies allows one to
build a semntic network of concepts that links molecules to other
molecules as well as other objects such as targets, genes and
pathways. How does one evaluate similarities in such a heterogenous
network \cite{couto2010}? How does one integrate chemical structure
information with ontologies \cite{hastingsowled2010}? And more
generally, how can we achieve automated inferencing in such
heterogenous linked systems, allowing us to acheive a systems level
understanding of small molecules \cite{Oprea:2007fk}.

@@

While the complexity of molecular graph enumeration remains an open
problem, an alternative to enumerating the space is to sample it
\cite{goldberg1999}. Sampling procedures based on the Metropolis or
the Genetic Algorithms have been developed and used to elucidate
natural compounds from NMR data. It should be noted that even in the
face of the computational complexity of molecular enumeration,
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB} (but only
considering molecules upto 13 heavy atoms and using only C, N, O, S
and Cl). Yet, the current enumeration software products do not
generally produce stereoisomers or tautomers; these require specific
enumeration procedures, which are still the subject of investigations
by the cheminformatics community.

Another relevant application of enumeration methods is in the
\emph{generation of chemical reaction networks}. The problem here
consists of enumerating all the possible compounds that can be
produced by applying reaction rules to a set of initial molecules. By
reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target, known as
\emph{retrosynthesis}. Designing new drugs or chemicals, understanding
the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and
synthetic biology to produce heterologous compounds in microorganisms,
all involve enumeration of reaction networks. As reviewed in Chapter
11 in \cite{faulon2010} several network enumeration techniques have
been developed. However, these techniques generally suffer from a
combinatorial explosion of product compounds. One way to limit the
number of compounds generated is to simulate the dynamics of the
network while it is being constructed and remove compounds of low
concentration. Following that idea, methods have been developed based
on the Gillespie Stochastic Simulation Algorithm (SSA) to compute
on-the-fly species concentrations. Chemical reaction network
enumeration and sampling is an active field of research, particularly
in the context of metabolism, either to study biodegradation, or to
propose metabolic engineering strategies to biosynthesize compounds of
commercial interest. The difficulty with metabolic network design is
that in addition to network generation based on reactions, one also
needs to verify that there are enzymatic events possible for enabling
reactions catalyses. That additional task requires encompassing both
chemical structure and protein sequence information and the
development of tools that are at the interface between cheminformatics
and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similar property principle}
\cite{Johnson:1990qf} (structurally similar molecules will exhibit
similar properties), the goal of any modeling approach is to capture
and characterize the correlations between structural features and the
observed biological activities. Simultaneously, such approaches must
also describe the likelihood of error when using such
models for decision making.  A variety of approaches can be employed
to assess the error in (or conversely, the reliability or confidence of) a
prediction, ranging from statistical approaches (XXX) to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input (i.e., chemical structures) that can be
reliably predicted, usually deliniated using some form of similarity
to the training set.

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
  Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking, pharmacophore modeling and proteochemometric methods
have been designed to take into account both ligand and receptor
structures. The last method is an example of an extension of
statistical QSAR methods to simultaneous model the receptor and
ligand as first reported by Lapinsh \textit{et  al.}~\cite{lapinsh2001}.

The first step in the predictive modeling of biological activities is
to generate molecular descriptors (a.k.a, features) that are numerical
representations of different structural features. Here multiple
approaches exist. We can transform molecular graphs into a numerical
vector representation (a.k.a. \emph{descriptors} or \emph{features})
of chemical structures \cite{todeschini2000}, ranging from counts of
elements to eigenvalues of the Laplacian matrix.  Alternatively, we
can compare molecular graphs directly, via kernel methods. In these
cases, rather than compute vectorial representations, we directly
operate on the graph representations.

Both methods have advantages and disadvantages. The vectorial approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors. This is the feature selection problem and
is well discussed in the data mining literature. A kernel approach
does not require feature selection; but now, one faces the problem of
evaluating a data set in a pairwise fashion. Furthermore, one must now
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy Mercer's condition, and this is
not always possible with traditional cheminformatics based kernels
(such as a multiple common substructure based kernel). These
challenges can make kernel based methods prohibitive on very large
datasets.

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we
looking for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
some of explanation as to why a molecule is predicted as toxic or
non-toxic, for example. The former scenario is common in virtual
screening settings, where we require high degree of accuracy and fast
predictions, but do not really care as to why one molecule is active
and another inactive. These types of models can be black boxes (neural
networks) and algorithmic models \cite{Breiman:2001fk} (random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we require to understand the effect of a
certain structural feature on the observed potency and expect that the
model will provide insight. These types of models are generally
distributional in nature (such as linear regression and PLS) though
some algorithmic approaches such as random forests can provide some
degree of insight.

Having chosen a modeling approach, a vital question is that of model
reliability which is closely tied to the concept of model
applicability -- the question of when the prediction of the model for
a new object is reliable. This issue has grown in importance with the
increasing use of predictive models in regulatory settings. A
misprediction (due to the model not bein applicable to a certain class
of inputs) can have significant financial repurcussions. A variety of
methods have been developed that attempt to characterize the domain of
a model. These methods not only determine whether models are
applicable, but can also be used for deciding if additional biological
experiments are required for reducing the prediction error on certain
compound classes.

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in
isolation. Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that such aspects be taking into account and, as
far as possible, take into account the receptor
simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Recent techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.

With the advent of high-throughput screening technologies, large
libraries of compounds can be screened against multiple targets in an
efficient manner. Such panel assays provide a broad, systems-level
view of small molecule activities.  Models developed on such data
afford us the opportunity to identify targets, characterize off-target
effects and so on. However, most approaches to this problem temd tp
develop multiple individual models \cite{Chen:2010zr}, leading to
multiple, independent predictions for a given input molecule.
Instead, one could imagine an approach that takes into account the
covariance structure of the multiple observed activities and the
structural descriptors within a single model. Such an approach could
lead to more robust predictions for panel assays. Finally, this
approach also better reflects clinically relevant compound
profiles~\cite{kuhn2010} and the ``personalized medicine'' concept
(\emph{e.g.}, drug-drug interaction profiles).

@@

The relevance of scientific knowledge management is increasing not
only for reducing the risk in current research, but also for enabling
new collaboration/innovation opportunities with internal partners and
a rapidly growing number of external/public partners. In
cheminformatics and chemistry, scientists already switched many years
ago from paper lab-notebooks to online collaboration platforms, called
ELNs (Electronic Lab Notebooks). So, what were the drivers behind
chemists, and companies, adopting ``Enterprise 2.0,'' a social online
collaboration culture?

@@

While ELNs technically existed as long ago as the 1990s, chemists
adopted their use in large numbers in 2004 and 2005, and the market is
still growing~\cite{ELNstatus}. As one report notes, ``The initial
drive for the development of ELNs came from the field of chemistry,
perhaps driven by the early adoption by chemists of computer
technologies \ldots The pharmaceutical industry shifted from a
position of `if we get an ELN' to `when we get an
ELN'.''~\cite{ELNstatus}. The growth of the use of ELNs has been
driven by many factors including ease of use (rapid searches across
organization-wide data sources, easy sharing of domain knowledge) and
regulatory compliance (timestamped experimental data, experiment and
data approvals and QC information). Given that any experiment could
become the core of a patent filing, ELNs allow organizations to define
and implement audit trails in a standardized and efficient fashion.

>>>>

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: The
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce the
  new therapeutic molecules.  Only within the last decade -- and as an
  effect of a change in mindsets, government-mandated policy changes,
  and, not least, because of chemists volunteering their time
  for an Open Science ``movement'' -- have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Making chemical information public is important because the drugs we
take are the result of intensive and costly research, which is
facilitated by the amount of information available. The more
information is shared (a process very hard to achieve in
pharmaceutical companies), the more information is available to each
researcher, facilitating the development of novel treatments. The
relevance of computer science to this research becomes clear when we
consider the quantity and types of data available -- currently a
single data base such as PubChem contains more than 34 million
chemical structure records, along with an even larger number of
\emph{annotations} (comments on the structure that are database searchable), such as synonyms (different names for the same molecule), known targets (drugs with which the molecule is known to interact), mode of action (how the molecule reacts with others),
regulatory approval status and so on.  Simply considering the synonym
list, PubChem stores 50.1 million synonym for 19.6 million compounds.

@@

Cheminformatics comprises different areas, which broadly speaking can
be divided into three fields: \emph{capturing data} (using lab
notebooks or potentially using formats such as the Chemical Markup
Language for publications); \emph{storing data} (designing database
schemas, devising ontologies) and \emph{mining data} (such as for
bioactivity prediction, which might involve algorithms based on
graphs).  However, the type of information being analyzed is different
from biological information: while bioinformatics often deals with
sequences, the domain of cheminformatics is chemical structures. In
the case of the former, information can very frequently be represented
as one-dimensional strings, which are relatively easy to handle
computationally. In the case of the latter however, chemical
structures may possess rings, branches as well as multiple valid
representations of the same structure (such as \emph{tautomers}, where hydrogen atoms can be positioned in different
places of the molecules), potentially giving rise to
ambiguities. Hence, chemical structures are often reported to be more
difficult to standardize.

@@

Although graphs implicitly encode the 3D structure of a molecule
(when combined with knowledge about favored bond angles and
distances), many different low energy 3D structures, or conformers,
may be consistent with the same graph structure. In addition, the 3D
structure may have a ``secondary'' geometrical arrangement of features
(such as the presence of a right-handed helix) which cannnot be
encoded in the graph. Thus we have 3D representations
(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
goes further and represents a molecular surface, encoding some
property (such as lipophilicity, the ability to dissolve in fats) that varies over the surface.

@@

There is considerable progress in open chemical and bioactivity
databases like ChEMBL and NCBI-PubChem, which are both freely
available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Still, the integration of
chemical databases with one other remains a challenge, not only due
to the normalization problems in the chemical and bioactivity
landscape, but also due to sheer data volume. One problem is the
annotation with the ``protein targets'' a molecule binds to -- here
dozens of protein and gene identifiers exist, generally without the
availability of a one-to-one mapping from each ontology, which makes
merging databases often a cumbersome exercise.

Given the requirement to ``make sense'' of as much chemical data at
hand as possible, we have to be able to search in complex data, for
example multi-labelled molecular graphs, where graph labels can change from database to database. Hence, we need to be able to capture molecular graphs in a
risk-bounded machine encoded way.

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. When one considers the need to include other aspects such as
protonation states (how the molecule changes when one proton is added to it) and tautomer states (how the molecule can change when a proton migrates from one part of the molecule to another), the challenge increases. The need for
a unique representation that is invariant with respect to the atom
ordering arises due to the fact that graph isomorphism (i.e., checking
whether two molecules are the same) is an expensive operation. The
first such ``unique representation algorithm,'' or \emph{canonicalization} algorithm, was described by Morgan
\cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format \cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

With each database using their own algorithm for molecular encoding,
automated exchange of data between different databases was hindered.
As more and more data became available online, the use of unique
structure-based identifiers became a pressing need, resulting in the
IUPAC International Chemical Identifier (InChI) to be developed. The
InChI is a non-proprietary, structured textual identifier for chemical
entities~\cite{inchi}. InChI identifiers are not intended to be read
and understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. Figure~\ref{figure:smiles} illustrates
the SMILES, InChI and InChIKey for lipoic acid.

@@

The InChI is now widely used for matching identical chemical
structures, but it is still limited. For example, it cannot
differentiate between certain types of stereoisomers (informally, molecules that are ``3D mirror images'' of each other):
Figure~\ref{figure:cistrans} (in the Appendix) illustrates two
stereoisomers for which the generated InChI is the same.

@@

Graph substructure matching is a variant of \emph{graph isomorphism},
which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is infeasible \cite{Weininger:2011ly}. Speedups can be
obtained via the use of heuristics such as structural
\emph{fingerprints}. Fingerprints encode characteristic features of a
given chemical structure, usually in a fixed-length
bitmap. Fingerprints fall broadly into two categories: structure keys
and hashed keys. In the former, each bit position corresponds to a
distinct substructure such as a \emph{functional group} (a connected set of atoms that affects the characteristics of the chemical reactions of the molecule). Examples
include MACCS and PubChem keys. The latter class are designed such
that substructural patterns are represented as strings and then hashed
to a random bit position. As a result, a given position can encode
multiple substructures. The advantage of such fingerprints is that
they can cover an arbitrarily large collection of substructures such
as ``paths of length $N$,'' or circular environments, and so on. Examples
include ECFPs (Extended Connectivity Fingerprints, which use local topological information) and Daylight fingerprints (which are ``folded'' to optimize information density and screening speed).

Given a binary fingerprint, we can first ``pre-screen'' a database, to
ignore molecules that cannot possibly match the query. This is
achieved by requiring that all bits in a query fingerprint must also
be present in the target fingerprint. Since the target fingerprints
are pre-computed, comparing binary strings can be performed extremely
rapidly on modern hardware. As a result, we apply the actual
isomorphism test only on those molecules that pass the
screen. Fingerprints can also be used to rapidly search databases for
similar molecules, in which case a similarity metric such as the
Tanimoto coefficient is used to compare the query and target
fingerprints. While fast, heuristics have been developed
\cite{Swamidass:2007ve} to further
speed up similarity searches.

@@

The preceding section focuses on the choice of representation of
chemical structures. With the profusion of databases that contain
details of small molecules, an emerging problem is the integration of
such databases. Such integration forms the basis of ``systems'' level
investigations of biological systems. A current challenge in
cheminformatics is how to identify ``equivalent'' objects across data
sources. At one level, equivalent compounds can be identified via
canonical structure representations (InChI). But there are other
equivalences beyond structure. For example, compounds may bind to the
same or similar targets; they may be have similar biological function
(say, anti-neoplastic, that is to say, it prevents the development of tumors) and so on. To assess such higher level
characteristics, ontologies and semantic technologies have been
employed, to first, support annotations of chemical systems and
second, enable semantic analyses of chemical systems in conjunction
with other biological systems (such as pathways). A number of
challenges remain in this area - the use of ontologies allows one to
build a semantic network of concepts that links molecules to other
molecules as well as other objects such as targets, genes and
pathways. Open questions include the following.  How does one evaluate similarities in such a heterogenous
network \cite{couto2010}? How does one integrate chemical structure
information with ontologies \cite{hastingsowled2010}? And more
generally, how can we achieve automated inferencing in such
heterogenous linked systems, allowing us to acheive a systems level
understanding of small molecules \cite{Oprea:2007fk}.

@@

While the complexity of molecular graph enumeration remains an open
problem, an alternative to enumerating the space is to sample it
\cite{goldberg1999}. Sampling procedures based on the Metropolis or
the Genetic Algorithms have been developed and used to elucidate
natural compounds from NMR data. It should be noted that even in the
face of the computational complexity of molecular enumeration,
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB} (but only
considering molecules up to 13 heavy atoms and using only carbon, nitrogen, oxygen, sulfur
and chlorine). Yet, the current enumeration software products do not
generally produce stereoisomers or tautomers; these require specific
enumeration procedures, which are still the subject of investigations
by the cheminformatics community.

Another relevant application of enumeration methods is in the
\emph{generation of chemical reaction networks}. The problem here
consists of enumerating all the possible compounds that can be
produced by applying reaction rules to a set of initial molecules. By
reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target, known as
\emph{retrosynthesis}. Designing new drugs or chemicals, understanding
the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and
synthetic biology to produce heterologous compounds (compounds from different species) in microorganisms,
all involve enumeration of reaction networks. As reviewed in Chapter
11 in \cite{faulon2010} several network enumeration techniques have
been developed. However, these techniques generally suffer from a
combinatorial explosion of product compounds. One way to limit the
number of compounds generated is to simulate the dynamics of the
network while it is being constructed and remove compounds of low
concentration. Following that idea, methods have been developed based
on the Gillespie Stochastic Simulation Algorithm (SSA) to compute
on-the-fly species concentrations. Chemical reaction network
enumeration and sampling is an active field of research, particularly
in the context of metabolism, either to study biodegradation, or to
propose metabolic engineering strategies to biosynthesize compounds of
commercial interest. The difficulty with metabolic network design is
that in addition to network generation based on reactions, one also
needs to verify that there are enzymatic events possible for enabling
reactions catalyses (in other words, enzymes must be present to reduce the energy required for some reactions to take place). That additional task requires encompassing both
chemical structure and protein sequence information and the
development of tools that are at the interface between cheminformatics
and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similar property principle}
\cite{Johnson:1990qf} (structurally similar molecules will exhibit
similar properties), the goal of any modeling approach is to capture
and characterize the correlations between structural features and the
observed biological activities. Simultaneously, such approaches must
also describe the likelihood of error when using such
models for decision making.  A variety of approaches can be employed
to assess the error in (or conversely, the reliability or confidence of) a
prediction, ranging from statistical approaches to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input (i.e., chemical structures) that can be
reliably predicted, usually deliniated using some form of similarity
to the training set.

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
  Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking (which predicts how molecules fit together), pharmacophore modeling (which predicts the surface features of a molecule) and proteochemometric methods (which predict the interaction between small molecules and proteins)
have been designed to take into account both ligand and receptor
structures. The last method is an example of an extension of
statistical QSAR methods to simultaneous model the receptor and
ligand as first reported by Lapinsh \textit{et  al.}~\cite{lapinsh2001}.

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that are numerical
representations of different structural features. Here multiple
approaches exist. We can transform molecular graphs into a numerical
vector representation of chemical structures \cite{todeschini2000}, ranging from counts of
elements to eigenvalues of the Laplacian matrix.  Alternatively, we
can compare molecular graphs directly, via \emph{kernel methods}, where a kernel on graphs $G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a kernel on a single graph compares measures similarities between the graph's nodes. In these
cases, rather than compute vectorial representations, we directly
operate on the graph representations.

Both methods have advantages and disadvantages. The vectorial approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors. This is the feature selection problem and
is well discussed in the data mining literature. A kernel approach
does not require feature selection; but now, one faces the problem of
evaluating a data set in a pairwise fashion. Furthermore, one must now
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy \emph{Mercer's condition} (a well-known mathematical property in machine learning that makes a set of observations easier to make predictions about), and this is
not always possible with traditional cheminformatics based kernels
(such as a multiple common substructure based kernel). These
challenges can make kernel based methods prohibitive on very large
datasets.

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we
looking for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
also some explanation as to why a molecule is predicted as toxic or
non-toxic, for example. The former scenario is common in virtual
screening settings, where we require a high degree of accuracy and fast
predictions, but do not really care why one molecule is active
and another inactive. These types of models can be black boxes (neural
networks) and algorithmic models \cite{Breiman:2001fk} (random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we need to understand the effect of a
certain structural feature on the observed potency and expect that the
model will provide insight. These types of models are generally
distributional in nature (such as linear regression and Partial Least Squares) though
some algorithmic approaches such as random forests can provide insight.

Having chosen a modeling approach, a vital question is that of model
reliability which is closely tied to the concept of model
applicability -- the question of when the prediction of the model for
a new object is reliable. This issue has grown in importance with the
increasing use of predictive models in regulatory settings. A
misprediction (due to the model not being applicable to a certain class
of inputs) can have significant financial repercussions. A variety of
methods have been developed that attempt to characterize the domain of
a model. These methods not only determine whether models are
applicable, but can also be used for deciding if additional biological
experiments are required for reducing the prediction error on certain
compound classes.

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in
isolation. Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that the 3D geometries of that molecule be taking into account and, as
far as possible, take into account the receptor
simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Recent techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.

With the advent of high-throughput screening technologies, large
libraries of compounds can be screened against multiple targets in an
efficient manner. Such panel assays provide a broad, systems-level
view of small molecule activities.  Models developed on such data
afford us the opportunity to identify targets, characterize off-target
effects and so on. However, most approaches to this problem tend to
develop multiple individual models \cite{Chen:2010zr}, leading to
multiple, independent predictions for a given input molecule.
Instead, one could imagine an approach that takes into account the
covariance structure of the multiple observed activities and the
structural descriptors within a single model. Such an approach could
lead to more robust predictions for panel assays (that is, which battery of tests would be most useful to perform). Finally, this
approach also better reflects clinically relevant compound
profiles~\cite{kuhn2010} and the ``personalized medicine'' concept
(\emph{e.g.}, drug-drug interaction profiles).

@@

The relevance of scientific knowledge management is increasing not
only for reducing the risk in current research, but also for enabling
new collaboration/innovation opportunities with internal partners and
a rapidly growing number of external/public partners. In
cheminformatics and chemistry, scientists already switched years
ago from paper lab-notebooks to online collaboration platforms, called
ELNs (Electronic Lab Notebooks). So, what were the drivers behind
chemists, and companies, adopting ``Enterprise 2.0,'' a social online
collaboration culture?

@@

While ELNs technically existed as long ago as the 1990s, chemists
adopted their use in large numbers in 2004 and 2005, and the market is
still growing~\cite{ELNstatus}. As one report notes, ``The initial
drive for the development of ELNs came from the field of chemistry,
perhaps driven by the early adoption by chemists of computer
technologies \ldots The pharmaceutical industry shifted from a
position of `if we get an ELN' to `when we get an
ELN'.''~\cite{ELNstatus}. The growth of the use of ELNs has been
driven by many factors including ease of use (rapid searches across
organization-wide data sources, easy sharing of domain knowledge) and
regulatory compliance (timestamped experimental data, experiment and
data approvals and quality control information). Given that any experiment could
become the core of a patent filing, ELNs allow organizations to define
and implement audit trails in a standardized and efficient fashion.
